---
title: "UKDriverDeaths" 
subtitle: "Using prophet to forecast future data for the number of road casualties in Great Britain" 
author: "Mariam Ashabi Subair Bello" 
date: "13/03/2025" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
---
 
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

The primary goal of this project is to utilize the Prophet function to forecast future values for the monthly "UKDriverDeaths" dataset and analyze how Prophet generates and presents this forecasted data.


```{r}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("prophet")
library(prophet)

```

CRAN mirrors are servers around the world that host R packages and I used it because R didn't know which CRAN mirror it should use so it wasn't installing the prophet packages. The install.packages("prophet") command was used to install prophet.Then library was used to load the packages into R.

```{r}
install.packages("zoo")
library(zoo)
```

The install.packages("zoo") command was used to install zoo. Then library was used to load the packages into R.

```{r}
data("UKDriverDeaths")
?UKDriverDeaths
```

The function data was used to load the dataset called "UKDriverDeaths". "?UKDriverDeaths" was used to enquire about the data which will explain the dataset to us.

```{r}
deaths.dataframe = data.frame(
    ds = zoo::as.yearmon(time(UKDriverDeaths)),
    y = UKDriverDeaths
)
```

Next, we created a data frame with two columns: ds for time and y for number of deaths or seriously injured. Where y is the monthly totals of car drivers in Great Britain killed or seriously injured from Jan 1969- Dec 1984.

```{r}
fitted_model = prophet::prophet(deaths.dataframe)
```

The function prophet is used to fit the data.

```{r}
future_deaths_dataframe = make_future_dataframe(fitted_model,periods = 8,freq = "quarter")
```

The make_future_dataframe function uses the ds column, which contains the dates, to generate a new dataframe with additional dates for future predictions. By specifying the number of periods, it extends the timeline into the future. In this case, we can see the new dates for 1995-1996, broken down into quarters, which were automatically created by the function. This allows us to forecast data for those future time periods.

```{r}
predict_future_deaths = predict(fitted_model,future_deaths_dataframe) 
```

The predict function is used to create the forecast which is the data containing the forecast of the yhat column.

## 2.2 Displaying plots
```{r}
plot(fitted_model,predict_future_deaths)
```

The graph plots dates against monthly totals of car drivers in Great Britain killed or seriously injured. The blue line shows Prophet's predicted values, while the black line represents the original data. It reveals a decreasing trend over time, though the data is noisy and irregular.

```{r}
prophet_plot_components(fitted_model,predict_future_deaths)

```

Here we can get more plots that show the trend and yearly seasonality of the forecast.
The trend shows that there has been a decrease of car driver killed or injured in Greaat Britain by 700 deaths.

The yearly seasonality shows that there are more deaths around July and August compared to the other months.

```{r}
install.packages("astsa")
library(astsa)

```

The install.packages("astsa") function is to download the astsa package to be able to do the next part of the code and library to load the packages into the R session.

```{r}
model23 <-stats::decompose(UKDriverDeaths)
plot(model23)

```

The model23 is used for a time series decomposition into three main components, in this case random, seasonal and trend but as you can see there is an extra component called observed. Errors are spread around zero which makes it a good model.
The trend time series shows how the number of drivers killed or injured has been decreasing.


